---
title: "README"
output: md_document  
rmarkdown::github_document:
pandoc_args: --webtex
header-includes:
   - \usepackage{amsmath, amssymb, amsthm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bayesian Additive Regression Trees for Multivariate skewed responses

This repository contains code to implement the methodology described in the paper "Bayesian Additive Regression Trees for Multivariate Skewed Responses", by Um, Linero, Sinha and Bandyopadhyay (2022+, Statistics in Medicine)


This package uses the primary functions from [`SoftBART`](https://github.com/theodds/SoftBART) to incorporate the SoftBART model as a component.

## Installation

The packages can be installed with the `devtools` package:

```{r, eval = FALSE}
library(devtools) 
devtools::install_github(repo='Seungha-Um/skewBART') 
```

The package with the vignette's included can be installed with 

```{r, eval = FALSE}
devtools::install_github(repo='Seungha-Um/skewBART',build_vignettes = TRUE) 
```

and then accessed by running `browseVignettes("skewBART")` (to reproduce our results, one will need to increase the number of MCMC samples). Alternatively, 
Vignette is available at [Simulation](https://rpubs.com/sheom0808/926961) and  [GAAD dataset](https://rpubs.com/sheom0808/926959).

One of the vignette's replicates our analysis of the GAAD dataset, a subset of which has been included, while the other illustrates our methods on simulated data.


