---
title: "skewBART for real data analysis (GAAD)"
output: rmarkdown::html_vignette
author: Seungha Um
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{real_data_GAAD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction 

This vignette demonstrates how to fit the skewBART and multivariate skewBART
models using data from the Type-2 diabetic Gullah-speaking African Americans
(GAAD) study. This dataset is included in the package, and more details on this
application can be found in Um, Linero, Sinha, and Bandyopadhyay (2022+,
Statistics in Medicine). We will use a small number of MCMC iterations and trees
to reduce computation times; these should instead be set to (say) 200 trees and
5000 iterations.

We begin by loading the required packages:
```{r, eval = TRUE, include=TRUE}
## Load library
library(tidyverse) # Load the tidyverse, mainly for ggplot
library(kableExtra) # For fancy tables
library(skewBART) # Our package
library(zeallot) # For the %<-% operator, used when generating data
options(knitr.table.format = 'markdown')
```


# Univariate skewed response

There are two responses in the GAAD dataset; the mean clinical attachment level
(CAL) and the mean periodontal pocket depth (PPD). We begin with a univariate
analysis of the CAL response. The data can be accessed in the package by running
the commands
```{r}
data(GAAD)
Y <- GAAD$subCAL
X <- GAAD %>% dplyr::select(Age, Female, Bmi, Smoker, Hba1cfull)
```
As in the `Vignette` vignette (see `browseVignettes("skewBART")`), we first
create hyperparameters and set the MCMC options:
```{r}
hypers <- UHypers(X, Y, num_tree = 20)
opts <- UOpts(num_burn = 250, num_save = 250)
```
We will use the `skewBART_pd` function to make partial dependence plots in order
to examine the marginal effects of some groups of covariates. The following
specifies the values (in the form of a `data.frame`) to compute the marginal
effect at:
```{r}
x_grid <- expand.grid(Hba1cfull = seq(5, 16.4, length = 20), 
                      Smoker = c(0,1), Female = c(0,1))
```
We then fit the model using the `skewBART_pd` command:
```{r, warning=FALSE}
set.seed(77777)
fitted_skewbart <- skewBART_pd(X, Y, vars = c("Hba1cfull", "Smoker", "Female"), 
                               x_grid = x_grid,
                               hypers = hypers, opts = opts)
```
Later, we will also need the fit of the model to PD:
```{r}
set.seed(77777)
PD <- GAAD$subPD
hypers <- UHypers(X, PD, num_tree = 20)
fitted_skewbart_pd <- skewBART_pd(X, PD, vars = colnames(x_grid), 
                                  x_grid = x_grid, 
                                  hypers = hypers, opts = opts)
```

We can compare this fit with a model fit on the log scale (for simplicity, we
don't compute the partial dependence here):
```{r, warning = FALSE}
set.seed(77777)
hypers <- UHypers(X, log(Y), num_tree = 20)
fitted_skewbart_log <- skewBART(X, log(Y), X, hypers, opts)
```
We might then compare the two fits by LPML; for the original model, this is
```{r}
fitted_skewbart$loo$estimates["elpd_loo",1]
```
while for the log model we need to take into account the Jacobian of the log
transformation $\log(Y)$ to give a fair comparison:
```{r}
fitted_skewbart_log$loo$estimates["elpd_loo",1] - sum(log(Y))
```
We see that the model with the log transformation performs better overall.

We next examine the posterior of the skewness $\alpha$:
```{r}
hist(fitted_skewbart$alpha)
hist(fitted_skewbart_log$alpha)
```

We see that, even after using the log transformation, the data still suggests
that we should use a skew-normal model over a normal model for the error
($\alpha$ is concentrated away from $0$).

Next, we examine the partial dependence plots to see (i) how Hba1c affects
CAL and (ii) how this effect is moderated by gender and smoking. The fitted
model contains both the samples and summary statistics (i.e., the posterior mean
and the 2.5th and 97.5th percentiles) of the partial dependence function:
```{r}
head(fitted_skewbart$partial_dependence_samples)
head(fitted_skewbart$partial_dependence_summary)
```
Here, $\widehat f$ is the BART-modeled function while $\widehat Y$ is the
predicted value after accounting for the fact that the errors are not mean 0.
We use this to plot the partial dependence:
```{r}
ggplot(fitted_skewbart$partial_dependence_summary, 
      aes(x = Hba1cfull, y = y_hat_mean, color = interaction(Female, Smoker))) +
  geom_point() + geom_line() + theme_bw() + ylab("Predicted CAL") + 
  theme(legend.position = "bottom")
```

The estimated effects are not quite homogeneous, with female smokers starting
closer to female non-smokers for low values of Hba1c and ending up nearly
equal to male non-smokers for high values of Hba1c. Overall, males have higher
values of CAL.

Next, we display log-CAL against its predicted values:
```{r}
qplot(fitted_skewbart_log$y_hat_train_mean, log(Y)) +
  geom_abline(slope = 1, intercept = 0) + theme_bw() +
  xlab("Predicted log-CAL") + ylab("Observed log-CAL")
```

We see that the model does a good job capturing the relationship between log-CAL
and the predictors.



# Multivariate skewed response

Next, we fit the multivariate skewBART model to both the CAL and PPD. The
outcome `Y` is now a 2 by 2 matrix:
```{r}
X <- GAAD %>% select("Age", "Female", "Bmi", "Smoker", "Hba1cfull") %>% as.matrix()
Y <- GAAD %>% select(subCAL, subPD) %>% as.matrix()
```
We then build the hyperparameter/MCMC objects:
```{r}
hypers <- Hypers(X = X, Y = Y, num_tree = 20)
opts <- Opts(num_burn = 50, num_save = 50, num_print = 10)
```
We can fit the model, creating partial dependence plots as before, using the
`MultiskewBART_pd` function:
```{r, warning = FALSE}
set.seed(77777)
fitted_Multiskewbart <- MultiskewBART_pd(
  X = X, Y = Y, vars = colnames(x_grid), x_grid = x_grid, hypers = hypers, 
  opts = opts
)
```
For the purposes of model comparison we can again look at the LPML:
```{r}
fitted_Multiskewbart$loo
```
And lastly we can construct our partial dependence plots:
```{r}
head(fitted_Multiskewbart$partial_dependence_summary)
fitted_Multiskewbart$partial_dependence_summary %>%
  ggplot(aes(x = Hba1cfull, y = y_hat_mean, 
             color = interaction(Smoker, Female))) + 
  geom_line() + geom_point() + facet_wrap(.~outcome, scales = "free_y") + 
  theme_bw() +theme(legend.position = "bottom") + ylab("Prediction") +
  xlab("Hba1c")
```

For comparison, let's also look at the same plots using univariate fits
to the data.
```{r}
skew_pd_cal <- fitted_skewbart$partial_dependence_summary %>%
  mutate(outcome = "subCAL", method = "skewBART")
skew_pd_pd <- fitted_skewbart_pd$partial_dependence_summary %>%
  mutate(outcome = "subPD", method = "skewBART")
mskew_pd <- fitted_Multiskewbart$partial_dependence_summary %>% 
  mutate(method = "MultiskewBART")
skew_pd <- rbind(skew_pd_cal, skew_pd_pd, mskew_pd)

skew_pd %>% 
  ggplot(aes(x = Hba1cfull, y = y_hat_mean, 
             color = interaction(Smoker, Female))) + 
  geom_line() + geom_point() + facet_grid(outcome~method, scales = "free_y") + 
  theme_bw() +theme(legend.position = "bottom") + ylab("Prediction") +
  xlab("Hba1c")
```

The results for MultiskewBART and skewBART are similar, with skewBART being much
smoother. When a larger number of iterations are used to fit the MultiskewBART
model, the results are otherwise quite similar.
