---
title: "Bayesian Additive Regression Trees for Multivariate skewed responses"
author: Seungha Um
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Additive Regression Trees for Multivariate skewed responses}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction 

This vignette demonstrates how to fit the skewBART model using the `skewBART` package. The small MCMC iterations and tree numbers are used to reduce computation time; in practice, larger values (say, 200 trees and 5000 MCMC iterations) should be used. Our model is an extension of BART to accommodate univariate and multivariate skewed responses.

First, we load the requisite packages:

```{r, eval = TRUE, include=TRUE}
## Load library
library(tidyverse) # Load the tidyverse, mainly for ggplot
library(kableExtra) # For fancy tables
library(skewBART) # Our package
library(zeallot) # For the %<-% operator, used when generating data
library(tictoc)
library(loo)
options(knitr.table.format = 'markdown')
```

# Univariate skewed response

We first illustrate on a simple example due to Friedman which takes
$$
  f(x) = 10 \sin(\pi \, x_1 \, x_2) + 20 (x_3 - 0.5)^2 + 10 x_4 + 5 x_5.
$$
We set $Y_i = f(x_i) + \epsilon_i$ with $\epsilon_i$ have a skew-normal
distribution with scale $\sigma = 3$ and shape $\alpha = 1$. 

## Setup

The following code creates a function to simulate data from the model.
```{r, eval = TRUE, include=TRUE}
## Create a function for Friedman’s example
sim_fried <- function(N, P, alpha, sigma) {
  lambda <- alpha * sigma/sqrt(1+alpha^2)
  tau <- sigma/sqrt(1+alpha^2)
  X <- matrix(runif(N * P), nrow = N)
  mu <- 10 * sin(pi * X[,1] * X[,2]) + 20 * (X[,3] - 0.5)^2 + 10 * X[,4] + 5 * X[,5]
  Z <- abs(rnorm(N, mean=0, sd=1) )
  Y <- mu + lambda * Z + rnorm(N, mean=0, sd=sqrt(tau))
  EY <- mu + lambda * sqrt(2/pi)
  return(list(X = X, Y = Y, EY = EY, mu = mu, Z=Z, tau = tau, lambda = lambda))
}
```
We then create a training set and testing set:
```{r, eval = TRUE, include=TRUE}
## Traning dataset : n = 100 observations, P = 5 covariates, sigma = 1, alpha = 3 
set.seed(12345)
c(X,Y,EY,mu,Z,tau,lambda) %<-% sim_fried(100, 5, 1, 3)
c(test_X,test_Y,test_EY,test_mu,test_Z,test_tau,test_lambda)  %<-% sim_fried(50, 5, 1 ,3)
```

Next, we create an object containing the hyperparameters for the model using the
`UHypers` function (U is for univariate, as opposed to multivariate). This
function uses default hyperparameters unless specific hyperparameters are
provided by the user; see `?UHypers` for more details.

```{r, eval = TRUE, include=TRUE}
hypers <- UHypers(X, Y, num_tree = 20)
opts <- UOpts(num_burn = 100, num_save = 500)
```

The function `UOpts` produces a list of the parameters for running the Markov
chain for univariate skewBART model. For illustration we use only 250 burn-in
and save iterations; the defaults are 2500 for each. `UOpts` can also be used
to control some features of the chain, such as whether hyperparameters are
updated.

## Fitting the skewBART model

After creating the hyperparameters and options, we fit the data by running
```{r, eval = TRUE, include=TRUE}
## Fit the model
tic("skewBART")
fitted_skewbart <- skewBART(X = X, Y = Y, test_X = test_X, 
                            hypers = hypers, opts = opts)
toc()
```

Alternatively, we can use the function `skewBART_parallel` running multiple `skewBART` threads in parallel as provided by the `parallel` package. The option `cores` specifies the number of cores to employ in multi-threading. The `skewBART_parallel` function is only available when you are running R on Mac OS or Linux or other variants of Unix. 


```{r}
tic("skewBART in parallel")
## parallel::mcparallel/mccollect do not exist on windows
if(.Platform$OS.type =='unix'){
fitted_skewbart_parallel <- skewBART_parallel(X = X, Y = Y, test_X = test_X, 
                         hypers = hypers, opts = opts, cores = 5)
}
toc()
```

## Contents of the model fit

The fitted model is a list containing the following elements:

- `y_hat_train`: samples with `num_save` rows containing samples of the
  estimated mean for each observation in the training set.
- `y_hat_test`: same as `y_hat_train`, but for the testing set.
- `y_hat_train_mean`: the posterior mean of the predicted value of each
  observation on the training set; this includes both the prediction from the
  BART model and an offset due to the fact that the skew-normal distribution
  does not have mean 0.
- `y_hat_test_mean`: same as `y_hat_train_mean` but on the testing set.
- `f_hat_train` and `f_hat_test` are analogous to `y_hat_train` and `y_hat_test`, 
   but just give the value of the BART function - for BART these are the same
   quantity, but for skewBART we need to account for the fact that the errors
   do not have mean 0.
- `sigma`: posterior samples of parameter $\sigma$ in the skew-normal error.
- `tau`: posterior samples of parameter $\tau$ in the skew-normal error.
- `alpha`: posterior samples of parameter $\alpha$ in the skew-normal error.
- `lambda`: posterior samples of parameter $\lambda$ in the skew-normal error.
- `likelihood_mat`: matrix of log-likelihood values to calculate the log pseudo marginal likelihood (LPML)


The parameters $(\lambda, \tau)$ and $(\alpha, \sigma)$ give equivalent
descriptions of the skew-normal distribution, and are related by the fact that
$\lambda = \alpha\sigma / \sqrt{1 + \alpha^2}$ and $\tau = \sigma / \sqrt{1 +
\alpha^2}$; the MCMC is run on $(\lambda, \tau)$, but the skew-normal (with
location parameter $0$) is typically parameterized in terms of $(\alpha,\sigma)$
as
$$
  p(y) = \frac{2}{\sigma} \phi\left(\frac{y}{\sigma}\right) \, \Phi\left(\alpha \frac{y}{\sigma}\right).
$$
So, for example, the posterior mean of the skewness level $\alpha$ is
```{r, eval = TRUE, include=TRUE}
mean(fitted_skewbart$alpha)
```
while the posterior mean of the predicted values on the test data is
```{r, eval = TRUE, include=TRUE}
fitted_skewbart$y_hat_test_mean
```

Below, we assess the mixing of the model with a traceplot (left) and plot the
fitted versus actual values (right) of $f(x)$.

```{r, eval = TRUE, include=TRUE, fig.width = 7}
# create a dataframe for the estimated alpha
df <- data.frame(iteration = 1:length(fitted_skewbart$alpha), alpha=fitted_skewbart$alpha)
p1 <- ggplot(df) + geom_line(aes(iteration, alpha), color = "darkgreen", alpha=0.7) + theme_bw() 

# create a dataframe for the fitted values
df_fitted <- data.frame(fitted = fitted_skewbart$f_hat_test_mean, observed = test_mu)
p2 <- ggplot(df_fitted) + 
  geom_point(aes(fitted, observed), alpha=0.8, shape=2, color = "darkgreen") +
  geom_abline(intercept = 0, slope = 1, alpha=0.8) +
  theme_bw() 
library(ggpubr)
ggarrange(p1, p2)
```

In our manuscript, log pseudo marginal likelihood (LPML) is computed to evaluate the model performance. Using the `loo` package, the LPML can be conveniently computed after obtaining the MCMC samples from the joint posterior.


```{r}
loo(fitted_skewbart$likelihood_mat, r_eff = NA)
```

LPML is obtained by the estimate of expected log pointwise predictive density (elpd_loo). You may have the warning message regarding Pareto k diagnostic values. In the `loo` package, LPML is computed using Pareto smoothed importance sampling and reliability of estimates is evaluated using the shape parameter k. In practice, good performance has been observed for values of k up to 0.7. See help('pareto-k-diagnostic') for details in the `loo` package. Note that this vignette is demonstrated under the small MCMC iterations and tree numbers to reduce computation time. To get k > 0.7, larger values of MCMC iterateions and tree numbers should be used. 



Also, the root mean square error can be computed with the posterior mean of the predicted values;

```{r, eval = TRUE, include=TRUE}
rmse <- function(x,y) sqrt(mean((x-y)^2))
rmse(test_Y, fitted_skewbart$y_hat_test_mean)
```




# Multivariate skewed response 

We extend Friedman's example to the bivariate setting by taking
$$
  Y_{ij} = f(x_i) + \epsilon_{ij}
$$
where $\epsilon_i \equiv (\epsilon_{i1}, \epsilon_{i2})$ has a multivariate
skew-normal distribution. Equivalently, we assume that $\epsilon_{ij}$ can be
written as
$$
  \epsilon_{ij} = \lambda_j \, Z_{ij} + \varepsilon_{ij}
$$
where $(\varepsilon_{i1}, \varepsilon_{i2})^\top$ follows a multivariate normal
distribution with mean $0$ and covariance matrix $\Sigma$, while the $Z_{ij}$'s
are independent truncated standard normal random variables. We take $\lambda_1 =
1, \lambda_2 = 3$ and $\Sigma$ a correlation matrix with correlation $\rho =
0.5$.

## Setup

The following code creates a function for generating the data.

```{r, eval = TRUE}
## Create a function for Friedman example with multivariate framework
sim_data_multi <- function(N, P, lambda, tau, rho) {
  X <- matrix(runif(N * P), nrow = N)
  mu <- 10 * sin(pi * X[,1] * X[,2]) + 20 * (X[,3] - 0.5)^2 + 10 * X[,4] + 5 * X[,5] 
  Z <- cbind(lambda[1] * abs(rnorm(N)), lambda[2] * abs(rnorm(N)))
  Sigma <- matrix(c(tau[1], sqrt(tau[1]*tau[2])*rho, sqrt(tau[1]*tau[2])*rho, tau[2]), 2, 2)
  Err <- MASS::mvrnorm(n=N, mu=c(0,0), Sigma = Sigma)
  Y <- cbind(mu, mu) + Z + Err
  EY <- rbind(mu, mu) + lambda * sqrt(2/pi)
  return( list(X = X, Y = Y, EY=EY, mu = mu, lambda = lambda, tau=tau, Z= Z, Sigma = Sigma) )
}
```

We then create the training and testing sets.

```{r, eval = TRUE}
## Simulate dataset
## Traning dataset : n = 100 observations, P = 5 covariates, 
## lambda = (2,3), tau = c(1,1), rho = 0.5.
set.seed(12345)
c(X,Y,EY,mu,lambda,tau,Z,Sigma) %<-% sim_data_multi(100, 5, c(2,3), c(1,1), 0.5)
c(test_X,test_Y,test_EY,test_mu,test_lambda,test_tau,test_Z,test_Sigma) %<-%
  sim_data_multi(50, 5, c(2,3), c(1,1), 0.5)
```

The functions `Hypers` and `Opts` are the multivariate extensions of `UHypers`
and `UOpts`. We use these to set the hyperparameters and MCMC options.

```{r, eval = TRUE}
## Create a list of the hyperparameters of the model. 
hypers <- Hypers(X = X, Y = Y, num_tree = 20)
opts <- Opts(num_burn = 50, num_save = 100)
```

## Fitting the multi-skewBART model

The model is fit using the `MultiskewBART` function:
```{r, eval = TRUE, include=TRUE}
tic("MultiskewBART")
fitted_Multiskewbart <- MultiskewBART(X = X, Y = Y, test_X = test_X, hypers=hypers, opts=opts) 
toc()
```

Alternatively, we can use the function `MultiskewBART_parallel` running multiple `MultiskewBART` threads in parallel as provided by the `parallel`. The option `cores` specifies the number of cores to employ in multi-threading. The `MultiskewBART_parallel` function is only available when you are running R on Mac OS or Linux or other variants of Unix. 


```{r}
tic("MultiskewBART in parallel")
## parallel::mcparallel/mccollect do not exist on windows
if(.Platform$OS.type =='unix'){
fitted_Multi_parallel <- MultiskewBART_parallel(X = X, Y = Y, test_X = test_X, 
                                                hypers = hypers, opts = opts,
                                                cores = 5)
}
toc()
```

## Contents of the Model Fit

The model fit contains quantities analogous to a univariate skewBART fit, but
now the dimension of the objects is different. For example. `y_hat_train` is now
an $N \times 2 \times$ `num_save` array, $\Sigma$ is a $2 \times 2 \times$
`num_save` array, and so forth. We can get the posterior mean of $\lambda_1$ and
$\lambda_2$ by computing
```{r, eval = TRUE, include=TRUE}
mean(fitted_Multiskewbart$lambda[,1])
mean(fitted_Multiskewbart$lambda[,2])
```
while the posterior mean of the function fit to the testing data is given by
```{r, eval = TRUE, include=TRUE}
head(fitted_Multiskewbart$f_hat_test_mean)
```
As before, we can examine the mixing of the skewness parameters $(\lambda_1,
\lambda_2)$ by running
```{r, eval = TRUE, fig.width=7}
# create a dataframe for estimated skewness levels (lambda)
df <- data.frame(iteration = 1:length(fitted_Multiskewbart$lambda[,1]),
                 lambda = c(fitted_Multiskewbart$lambda[,1], fitted_Multiskewbart$lambda[,2]),
                 grp = rep(c("lam1","lam2"), each=length(fitted_Multiskewbart$lambda[,1])))

ggplot(df) + geom_line(aes(iteration, lambda, colour=grp), alpha=0.7) +
  theme_bw() + facet_wrap(~grp) +
  scale_colour_manual(values=c(lam1 = "darkgreen", lam2= "brown")) +
  theme(legend.title=element_blank())
```
and plot of the fitted vs actual values by running
```{r, eval = TRUE, fig.width=7}
df_fitted <- data.frame(fitted = c(fitted_Multiskewbart$f_hat_test_mean),
                        truth = rep(test_mu, 2), 
                        grp = rep(c("response_1", "response_2"),each = length(test_mu)))

ggplot(df_fitted) + geom_point(aes(fitted, truth, colour=grp), alpha=0.8, shape=2) +
  geom_abline(intercept = 0, slope = 1, alpha=0.8) +
  scale_colour_manual(values=c(response_1 = "darkgreen", response_2= "brown")) +
  theme_bw() + facet_wrap(~grp) + theme(legend.title=element_blank())
```

The log pseudo marginal likelihood (LPML) can be computed using `likelihood_mat`.

```{r}
loo(fitted_Multiskewbart$likelihood_mat, r_eff = NA)
```

Also, the root mean square error can be computed with the posterior mean of the predicted values;

```{r, eval = TRUE, include=TRUE}
rmse <- function(x,y) sqrt(mean((x-y)^2))
rmse(test_Y, fitted_Multiskewbart$y_hat_test_mean)
```


One may wish to use the multi-BART, the multivariate version of the standard BART model. In this case, set the option `do_skew` to FALSE.

```{r, eval = FALSE, include=TRUE}
fitted_Multiskewbart <- MultiskewBART(X = X, Y = Y, test_X = test_X, 
                                      do_skew = FALSE, hypers=hypers, opts=opts) 
```



